Optimistic Atomic Multicast in One Communication Delay

### 2011-07-25 textos/optpaxos/draft_dropbox.tex
alg description

\documentclass[times, 10pt]{article} 
%\documentclass[times, 10pt,twocolumn]{article} 
%\usepackage[noend]{distribalgo}
\usepackage{algorithm}
%\usepackage{times}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{color}
\usepackage{amssymb}
\usepackage{xparse}
\usepackage[noend]{distribalgo}
\usepackage[draft]{fixme}
\usepackage[hmargin=2.5cm,vmargin=2.2cm]{geometry}

@@

\NewDocumentCommand \mycast { g }
{\IfNoValueTF{#1}{\amcast}{\amcastarg{#1}}}

@@

Hereafter, we assume that in addition to a set of destination groups, $m.dst$, each message $m$ has a source group, $m.src$, and a timestamp, $m.ts$. To multicast $m$ process $p$ in $g$ sets $m.ts$ to a unique timestamp based on its real-time local clock and then r-mcasts $m$ to all processes in $g$.

When processes in $g$ r-deliver $m$, they run a consensus instance to agree on the final timestamp assigned to $m$, after possibly adjusting it to ensure the following invariant: for any two messages $m$ and $m'$, multicast by processes in $g$ and decided in consensus instances $l$ and $l'$, respectively, where $l < l'$, it follows that $m.ts < m'.ts$. This is important since we intend to deliver messages according to their timestamp order.

@@

Once group $g$ has decided on $m$'s final timestamp, $m$ is r-mcast to all its destination groups. Let $h$ be a group in $m.dst$. Process $q \in h$ will deliver $m$ as soon as it has received at least one message from every group $i$ in $sendersTo(h)$ with a timestamp greater than $m$'s timestamp.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:deliveryminimal}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $propose_g(k, v)$, which proposes a value $v$ for the consensus instance of id $k$ within its group; and also $decide_g(k, v)$, which is called when the consensus instance of id $k$ finishes. $decide_g(k, v)$ is called for all the processes of the group that initiated it, when they learn that the value $val$ has been agreed upon in instance of id $k$. For the sake of simplicity, we assume that, for consensus instances within the same group, the values are decided in the same order of the instances id's\footnote{This can be easily done by delaying the callback of $decide_g(k, v)$ while there is some unfinished consensus instance of id $k': k' < k$ from the same group.}. Finally, we also use a fifo reliable multicast primitive \rmcastarg{$m$}{$groupSet$}, which \rmdel{}s $m$ to all the processes in all the groups in $groupSet$ in one communication step, in fifo order (e.g. the one described in \cite{ufrmcast1delta}).

Moreover, each process $p$ of group $g$ keeps some sets of messages:
\begin{itemize}
  %\item \textit{optPending}, which contains the messages waiting to be \mbox{\textit{\opt{}ed}} (optimistically delivered);
  \item \textit{\localmsgs}, containing the %local messages, i.e., the 
  messages that have been \amcast{} by some process of group $g$;
  \item \textit{\decided}, which contains the messages that have already been proposed and decided within $group(p)$;
  \item \textit{\stamped}, containing the messages sent to $g$ by any group; a final timestamp has already been assigned to each of these messages, although, in order to be \amdel{}ed, some of them may still need barriers from the groups in $sendersTo(G)$;
  \item \textit{\delivered}, which contains the messages that have been already \amdel{}ed by $p$.
\end{itemize}

@@

Something that must be noticed is that each message $m$ might not have its source group as a destination. Anyway, $m$ still has to be agreed upon in its group of origin $G$, so that its order among other messages from $G$ may be decided and for %$m.gs$ can be decided, and also for 
$m$ to be retrievable even in the presence of failures.% Because of that, to ensure that messages of this kind are also received by $group(p)$'s coordinator $c$, which then initiates the Paxos instance that allows for the conservative delivery of the message in all of its destinations, every message $m$ from $p$ must also be sent to $c$ (l. \ref{algline:sendcoord}). This way, when the clock value at the coordinator $time(c) > w(c) + m.ts$, $c$ will conservatively send $m$ to all the processes in $m.dst$. Finally, to avoid delivering a message $m$ in a group it is not addressed to, before OPT-Delivering it, each process must check whether it is one of $m$'s destinations or not (l. \ref{algline:notinterested}). In l. \ref{algline:sendcoord}, $m$ is sent to the whole group $G$, so that in the case of failure of $G$'s coordinator, it is more likely that the new coordinator will know about the existence of $m$.

The total order delivery of messages in a group can be solved by using consensus. Each consensus instance agrees upon some message set as the next ones to be delivered -- messages within the same set are told apart by the timestamp $m.ts$ applied by the process which created them; to solve timestamp collisions, the unique id of the sender process can be used. If processes send messages to each other using fifo reliable channels, and after receival, messages are proposed via consensus in the order in which they are received, the fifo delivery order is ensured.

@@

Here, we prove that the \amcast{} primitive ensures the properties defined in section \ref{sec:model} -- uniform validity, uniform agreement, uniform integrity, uniform total order and fifo order. To do this, we prove that \mbox{Algorithm \ref{algorithm:deliveryminimal}}, when used along with \mbox{Algorithm \ref{algorithm:nullperiodic}}, ensures these properties.

@@

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:barrierperiodic}
  Given a message $m$ in the $\stamped$ set of a correct process $p$ from group $g$, $p$ will eventually receive some $barrier(h) > m.ts$ from every group $h \in sendersTo(g)$ and insert it into its $stamped$ set.
\end{lems}

\begin{proof}Each process $q$ from every group $h$ keeps executing Algorithm \ref{algorithm:nullperiodic}, thus sending ever-increasing barrier values to all processes that might receive a message from $h$, i.e., to every process $r$, such that $\exists i \in \Gamma: h \in sendersTo(i) \wedge r \in i$. %Also, if $G \in receiversFrom(G')$, then $G' \in sendersTo(G)$. 
This means that, at some point, all correct processes of each group $h$ in $sendersTo(g)$ will have inserted a message $m'$ -- be it $null$ or not -- with a timestamp greater than or equal to that of $m$ into their respective $\localmsgs$ lists (l. \ref{algline:enqnull} of Algorithm \ref{algorithm:nullperiodic}). From Lemma \ref{lemma:mcastdecided}, all processes of $h$ will decide $m'$ and \rmcast{} it to $g$. From the properties of the \rmcast{} primitive, $p$ will \rmdel{} $m'$ and insert it into its $\stamped$ set (l. \ref{algline:insbp1} of Algorithm \ref{algorithm:deliveryminimal}). From the algorithm, the timestamps of the messages no longer change after being inserted into $\stamped$ so, eventually, $p$ will have in such set a barrier with a value greater than or equal to $m.ts$ from each group in $sendersTo(g)$.\end{proof}

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:lowertsfrmcastfirst}
Given two messages $m$ and $m'$ sent by the same group $g$, if a process $p \in g$ decides that their final timestamps are such that $m.ts < m'.ts$, then $p$ does not \rmcast{} $m'$ before $m$.
\end{lems}

\begin{proof}
Assume, by way of contradiction, that $p$ \rmcast{}s $m'$ first. This means that $m'$ was inserted into $\decided$ first. When $m$ is handled in the algorithm (lines \ref{algline:decidedloopbegin} to \ref{algline:rmothers}), $m'$ is already there. Then, $p$ sets the timestamp of $m$ to a value $m.ts > m'.ts$ (lines \ref{algline:decidedloopbegin} to \ref{algline:changetsafterdecision}), which is a contradiction.
\end{proof}

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:agreetimestamps}
Given any two processes and any message $m$, if each of such processes is either a destination of $m$ or belongs to $m.src$, then such processes agree on the final timestamp of $m$.
\end{lems}

@@

\noindent Base case ($k=1$): As this is the first consensus instance of the group, it means that $\decided = \emptyset$. From the agreement property of consensus, we know that $p$ and $q$ decided the same contents for $msgSet$ \mbox{(l. \ref{algline:decide})}. Each message included in such agreed set also includes an initial timestamp field. As the $\decided$ set is empty, such timestamps are not changed in neither $p$ or $q$ (lines \ref{algline:decidedloopbegin} to \ref{algline:changetsafterdecision}). Therefore, both processes consider the same timestamp value for every message. Besides, as the $msgSet$ is the same for both, the $\decided$ set remains identical in both $p$ and $q$.\\

\noindent Induction step: Suppose that $p$ and $q$ have already learnt the decisions of instance $k$, their $\decided$ sets remained identical and they decided the same timestamp value of each message sent from some process in their group so far. From the algorithm, all processes within a group learn all decisions in the same order, which is that of the consensus instance identifiers. Therefore, both $p$ and $q$ will next learn the decision of the instance $k+1$. From the agreement property of consensus, the $msgSet$ decided is the same for both processes. In the Algorithm \ref{algorithm:deliveryminimal}, the timestamps may be changed after the consensus decision only in line \ref{algline:changetsafterdecision}, based on what is already in the $\decided$ set (lines \ref{algline:decidedloopbegin} to \ref{algline:changetsafterdecision}). As the $msgSet$ and $\decided$ sets are each identical in $p$ and $q$, they will make the exact same change to the timestamp of each message in the \textit{while} loop (lines \ref{algline:decidedloopbegin} to \ref{algline:rmothers}). Therefore, they also agree on the timestamps of each message decided on consensus instance $k+1$ and their $decided$ sets remain identical.\\

Now, consider the case of $q$ not belonging to $g$. Let $r$ be any process of such group. After setting the final timestamp of $m$ (l. \ref{algline:decidedloopbegin} to l. \ref{algline:rmothers}), $r$ \rmcast{}s $m$ to $group(q)$ (l. \ref{algline:rmothers}). From the algorithm, after \rmdel{}ing $m$, $q$ never changes the value of $m.ts$, set by $q$ in accordance with $p$.% As any two processes on the same group from which $m$ has been multicast agree on its timestamp, and since any process from other group which is also a destination of $m$ agree on its timestamp as well, than any two processes that are destinations of $m$ agree on its timestamp.
\end{proof}

@@

\begin{proof}
Consider Algorithm \ref{algorithm:deliveryminimal}. Every process $q$ from $h$ sends messages to the processes of $g$, including $p$, using a fifo channel, via the \rmcast{} primitive. From the agreement property of consensus, we know that all processes from $h$ will decide the same messages in each consensus instance $k$. From Lemma \ref{lemma:lowertsfrmcastfirst} and Lemma \ref{lemma:agreetimestamps}, every process from $h$ \rmcast{}s to $g$ every message $m : m.src = h \wedge g \in m.dst$ (l. \ref{algline:rmothers} of Algorithm \ref{algorithm:deliveryminimal}) in the same order of timestamps. From the fifo property of \rmcast{}, we know that $p$ will \rmdel{} all of such messages from each process of $h$ in ascending order of timestamps. Although such messages will arrive from different senders and may be out of order at $p$, once $p$ received some message $m$ from some process of $h$, $p$ has also received any message \mbox{$m' : m'.src = h \wedge g \in m.dst\wedge m'.ts < m.ts$}.
%
%Assume, by way of contradiction, that $p$ received a message $m' : m'.ts < m.ts$ from some process $r \in h$ after having received $m$ from $q$. As $m'.src = m.src$ and $q$ \rmcast{}s to $g$ every message $m : m.src = h \wedge g \in m.dst$, this means that $q$ will send $m'$ afterwards, contradicting the fact that the processes in $h$ agree on each batch of messages decided and send them to $g$ in the same fifo order.
\end{proof}

@@

\begin{proof}
Let $g$ be the group of $p$. Group $g$ is either $m.src$ or not. In the former case, beforing inserting any message $m' : m'.src = m.src$ into $\stamped$, $p$ has to decide $m'$ first. When a message is decided by $p$, $p$ checks whether some other message which was decided previously has a timestamp lower than $m'$ and, if that is the case, the timestamp of $m'$ is changed to a value greater than that of any other message already decided (lines \ref{algline:decidedloopbegin} to \ref{algline:changetsafterdecision} of Algorithm \ref{algorithm:deliveryminimal}). Only then $m'$ may be inserted into the $\stamped$ set of $p$ (l. \ref{algline:insbp2}). Therefore, $p$ inserts messages from other processes of $g$ into its $\stamped$ set in ascending order of timestamps. Besides, the messages from $g$ are \rmcast{} to other groups in this same order (l. \ref{algline:rmothers}) by all correct processes of $g$, from the uniform agreement property of consensus.

@@

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:oncebpwilldel}
Once a message $m \neq null$ has been inserted into the $\stamped$ set of a correct process $p$, $m$ will eventually be delivered by $p$.
\end{lems}

\begin{proof}
Eventually, $p$, which necessarily belongs to some group $g$, will have received a barrier from every group in $sendersTo(g)$ with a value greater than the timestamp of $m$ (from Lemma \ref{lemma:barrierperiodic}). Once this happens, as any barrier is also a message, no more messages with a timestamp lower than that of $m$ will be inserted into the $\stamped$ set of $p$ (from the definition of $sendersTo(g)$ and Lemma \ref{lemma:oncedecidednotsmaller}). Let $ready$ be the set of $m$ plus all undelivered messages from $\stamped$ whose timestamps are lower than that of $m$, that is, $ready = \{m\} \cup \{m' : m' \in \stamped \setminus \delivered \wedge m'.ts < m.ts\}$.

We can infer that no more messages will be included in this set, meaning that any other message that might be later inserted into $\stamped$ will have a timestamp greater than that of any message in $ready$. Besides, the barrier received for $m$ also applies to all the other messages in this set, since their timestamps are lower than that of $m$. We can prove that each one of these messages will eventually be inserted into $\delivered$ and, if it is different from $null$, it will be delivered by $p$. We prove it by induction on the position $i$ of each message $m_i$ in $ready$, in ascending order of timestamps.\\
\\
\noindent Base case ($i=1$): Let $m_1$ be the first message in $ready$, i.e., \mbox{$\nexists m \in \stamped \setminus \delivered : m.ts < m_1.ts$}. We know that all the necessary barriers have been received already, so $m_1$ satisfies all conditions from lines \ref{algline:condel1} and \ref{algline:condel2} of \mbox{Algorithm \ref{algorithm:deliveryminimal}}, being inserted into the $\delivered$ set, after which $m_1$ no longer belongs to $\stamped \setminus \delivered$ and, if $m_1 \neq null$, $m_1$ will be delivered.\\
\\
\noindent Induction step: Suppose that $m_{i}$ will eventually be inserted into the $\delivered$ set and, if $m_i \neq null$, it will be delivered. Once this happens, it means that $m_{i}$ was the first message in $\stamped \setminus \delivered$ in ascending timestamp order. Since no more messages have been inserted into $ready$, as soon as $m_{i}$ is inserted into $\delivered$, $m_{i+1}$ will be the first one in $\stamped \setminus \delivered$, having, therefore, the lowest timestamp in such set. Then, as all barriers necessary for $m$ have already been received and $m_{i+1}.ts \leq m.ts$, $m$ will satisfy both conditions of lines \ref{algline:condel1} and \ref{algline:condel2} of \mbox{Algorithm \ref{algorithm:deliveryminimal}}. Thus, $m_{i+1}$ will also be inserted into the $\delivered$ set and, if $m_{i+1} \neq null$, it will be delivered.
\end{proof}

@@

\begin{proof}
From lines \ref{algline:condel1} to \ref{algline:addtodelivered}, and since no message is removed from The $delivered$ set, no message can be delivered twice, satisfying (3). For a message $m$ to be delivered (l. \ref{algline:consdeliver}), it must belong to the $\stamped$ set (l. \ref{algline:condel1}).  There are two possibilities to when $m$ has been inserted into such set:
\begin{itemize}
  \item $m$ has been originated in the same same group of $p$, that is, $p \in m.dst$, which means $m$ was inserted into $\stamped$ in line \ref{algline:insbp2} of Algorithm \ref{algorithm:deliveryminimal}, or
  \item $m$ has been sent from a group $h \neq g$, which means that it was inserted by $p$ into $\stamped$ in line \ref{algline:insbp1}. 
\end{itemize}

@@

As for the case when $p \notin m.src$, $m$ is inserted into the $stamped$ set of process $p$ in line \ref{algline:insbp1}, when it has been \rmdel{}ed. For a process $q \in h$, such that $h \neq g$, to \rmcast{} $m$, $m$ must have been decided within $h$, which means that it was proposed within $h$, therefore it was \amcast{} -- satisfying condition (1) -- by $h$ to $g$, which, from line \ref{algline:rmothers}, necessarily belongs to $m.dst$, satisfying condition (2).

@@

\begin{proof}
Consider Algorithm \ref{algorithm:deliveryminimal}. Suppose, by way of contradiction, that $m'$ has already been delivered by $p$, whereas $m$ has not. Message $m$ might have already been inserted into $\stamped$ or not. In the former case, as $m \neq null$ and $m$ has not been delivered, then $m$ belongs to $\stamped \setminus \delivered$. Therefore, $m'$ could not have been delivered, since $m.ts < m'.ts$ and it would not satisfy the condition from line \ref{algline:condel1} until $m$ has been delivered, so we have a contradiction in the case where $m$ was already in $\stamped$.

>>>>

\documentclass[times, 10pt]{article} 
%\documentclass[times, 10pt,twocolumn]{article} 
%\usepackage[noend]{distribalgo}
\usepackage{algorithm}
%\usepackage{times}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{color}
\usepackage{amssymb}
\usepackage[noend]{distribalgo}
\usepackage[draft]{fixme}
\usepackage[hmargin=2.5cm,vmargin=2.2cm]{geometry}

@@

%\NewDocumentCommand \mycast { g }
%{\IfNoValueTF{#1}{\amcast}{\amcastarg{#1}}}

@@

Hereafter, we assume that in addition to a set of destination groups, $m.dst$, each message $m$ has a source group, $m.src$, and a timestamp, $m.ts$. To \amcast{} $m$, process $p$ in group $g$ sets $m.ts$ to a unique timestamp based on its real-time local clock and then \rmcast{}s $m$ to all processes in $g$.

When processes in $g$ \rmdel{} $m$, they run a consensus instance to agree on the final timestamp assigned to $m$, after possibly adjusting it to ensure the following invariant: for any two messages $m$ and $m'$, \amcast{} by processes in $g$, if a process in $g$ decides $m$ before $m'$, then $m.ts < m'.ts$. 
% and decided in consensus instances $l$ and $l'$, respectively, where $l < l'$, it follows that $m.ts < m'.ts$. 
This is important since we intend to \amdel{} messages according to their timestamp order.

@@

Once group $g$ has decided on $m$'s final timestamp, $m$ is \rmcast{} to all its destination groups. Let $h$ be a group in $m.dst$. To ensure that no message $m'$, where $m'.ts < m.ts$, is \amdel{}ed after $m$, process $q \in h$ will \amdel{} $m$ only when it has received at least one message from every group $i$ in $sendersTo(h)$ with a timestamp greater than $m$'s timestamp.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:deliveryminimal}. We use a $getTime()$ primitive, which returns the current value of the local wallclock, which is monotonically increasing. Moreover, each process $p$ of group $g$ keeps some sets of messages:
\begin{itemize}
  %\item \textit{optPending}, which contains the messages waiting to be \mbox{\textit{\opt{}ed}} (optimistically delivered);
  \item \textit{\localmsgs}, containing the %local messages, i.e., the 
  messages that have been \amcast{} by some process of group $g$;
  \item \textit{\decided}, which contains the messages that have already been proposed and decided within $group(p)$;
  \item \textit{\stamped}, containing the messages sent to $g$ by any group; a final timestamp has already been assigned to each of these messages, although, in order to be \amdel{}ed, some of them may still need barriers from the groups in $sendersTo(G)$;
  \item \textit{\delivered}, which contains the messages that have been already \amdel{}ed by $p$.
\end{itemize}

@@

Even though each message $m$ might not have its source group as a destination, $m$ is still agreed upon in its group of origin, so its order among other messages from $m.src$ is decided and it %$m.gs$ can be decided, and also for 
becomes retrievable even in the presence of failures.% Because of that, to ensure that messages of this kind are also received by $group(p)$'s coordinator $c$, which then initiates the Paxos instance that allows for the conservative delivery of the message in all of its destinations, every message $m$ from $p$ must also be sent to $c$ (l. \ref{algline:sendcoord}). This way, when the clock value at the coordinator $time(c) > w(c) + m.ts$, $c$ will conservatively send $m$ to all the processes in $m.dst$. Finally, to avoid delivering a message $m$ in a group it is not addressed to, before OPT-Delivering it, each process must check whether it is one of $m$'s destinations or not (l. \ref{algline:notinterested}). In l. \ref{algline:sendcoord}, $m$ is sent to the whole group $G$, so that in the case of failure of $G$'s coordinator, it is more likely that the new coordinator will know about the existence of $m$.

Whenever $p$ has some message $m$ to \amcast, it assigns to $m$ an initial timestamp value, based on the local wallclock of $p$, after which it \rmcast{}s $m$ to all the other processes of $g$ (lines \ref{algline:gettime} and \ref{algline:rmlocal}). After \rmdel{}ing $m$, each process puts it in $\localmsgs$ to be proposed in $g$ until it is decided. As all correct processes of $g$ do so, some accepted proposal will end up containing $m$.

The consensus instances within $g$ are decided in the same order for every correct process of $g$. Such order is used to define the delivery order of $g$'s messages as well. We want this order to be consistent with the timestamps of the messages, so such timestamps can be changed if they are not in accordance with the delivery order.

Each consensus instances may decide not only one, but a batch of messages. When a message set is decided, the messages in it are handled in ascending order of timestamps, as if they were decided separately in the original timestamp order.

The atomic delivery of messages in a group can be solved by using consensus. Each consensus instance decides some message set as the next ones to be delivered -- messages within the same set are told apart by the timestamp $m.ts$ applied by the process which created them; to solve timestamp collisions, the unique id of the sender process can be used. If processes send messages to each other using fifo reliable channels, and after receival, messages are proposed via consensus in the order in which they are received, the fifo delivery order is ensured.

@@

Here, we prove that the \amcast{} primitive ensures the properties defined in section \ref{sec:model} -- uniform validity, uniform agreement, uniform integrity, uniform total order and fifo order. To do this, we prove that \mbox{Algorithm \ref{algorithm:deliveryminimal}}, when used along with \mbox{Algorithm \ref{algorithm:nullperiodic}}, ensures these properties. Although there are different processes executing such algorithms at the same time, we assume that there is no concurrency on the execution of the algorithms in any single process.

@@

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:barrierperiodic}
  Given a correct process $p$ and a message $m$, $p$ eventually receives some message $m' : m'.ts > m.ts$ from every group in $sendersTo(gropu(p))$.% and insert it into its $stamped$ set.
\end{lems}

\begin{proof}Let $g = group(p)$ and let $h$ be any group in $sendersTo(g)$. Every process $q \in h$ executes Algorithm \ref{algorithm:nullperiodic}, inserting messages with ever-increasing timestamp values into $\localmsgs$. Therefore, at some point, every process from $h$ will propose some message $m' : m'.ts > m.ts$. As one of these proposals will be decided, then one of these messages will be \rmcast{} to $g$. From the properties of \rmcast, $g$ will deliver it.
%
%which means that, from Lemma \ref{lemma:mcastdecided}
% 
% 
% thus inserting messages with ever-increasing timestamp values to all other groups that might receive a message from $h$, i.e., to every process $r$, such that $\exists i \in \Gamma: h \in sendersTo(i) \wedge r \in i$. %Also, if $G \in receiversFrom(G')$, then $G' \in sendersTo(G)$. 
% Therefore, at some point, all correct processes of each group $h$ in $sendersTo(g)$ will have inserted a message $m'$ -- be it $null$ or not -- with a timestamp greater than or equal to that of $m$ into their respective $\localmsgs$ lists (l. \ref{algline:enqnull} of Algorithm \ref{algorithm:nullperiodic}). From Lemma \ref{lemma:mcastdecided}, all processes of $h$ will decide $m'$ and \rmcast{} it to $g$. From the properties of the \rmcast{} primitive, $p$ will \rmdel{} $m'$ and insert it into its $\stamped$ set (l. \ref{algline:insbp1} of Algorithm \ref{algorithm:deliveryminimal}). From the algorithm, the timestamps of the messages no longer change after being inserted into $\stamped$ so, eventually, $p$ will have in such set a barrier with a value greater than or equal to $m.ts$ from each group in $sendersTo(g)$.
\end{proof}

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:lowertsfrmcastfirst}
Given any two messages $m$ and $m'$ sent from the same group $g$, if a process $p \in g$ decides that their final timestamps are such that $m.ts < m'.ts$, then $p$ does not \rmcast{} $m'$ before $m$.
\end{lems}

\begin{proof}
Assume, by way of contradiction, that $p$ \rmcast{}s $m'$ first. This means that $m'$ was inserted into $\decided$ first. When $m$ is handled in the algorithm (lines \ref{algline:decidedloopbegin} to \ref{algline:rmothers}), $m'$ is already there. Then, $p$ decides that the final timestamp of $m$ is $m.ts > m'.ts$ (lines \ref{algline:decidedloopbegin} to \ref{algline:changetsafterdecision}), which is a contradiction.
\end{proof}

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:agreetimestamps}
Given any two correct processes and a message $m$, if each of such processes is either a destination of $m$ or belongs to $m.src$, then such processes agree on the final timestamp of $m$.
\end{lems}

@@

\noindent Base case ($k=1$): As this is the first consensus instance within $g$, it means that $\decided$ was empty when such instance started. From the agreement property of consensus, we know that $p$ and $q$ decide the same contents for $msgSet$ \mbox{(l. \ref{algline:decide})}. Each message included in such agreed set also includes an initial timestamp field. As the $\decided$ set is empty, such timestamps are not changed in neither $p$ or $q$ (lines \ref{algline:decidedloopbegin} to \ref{algline:changetsafterdecision}). Therefore, both processes consider the same timestamp value for every message. Besides, as the $msgSet$ is the same for both, the $\decided$ set remains identical in $p$ and $q$.\\

\noindent Induction step: Suppose that $p$ and $q$ have already learnt the decisions of instance $k$, their $\decided$ sets remained identical and they decided the same timestamp value for each message sent from some process in their group so far. From the algorithm, all processes within a group learn all decisions in the same order, which is that of the consensus instance identifiers. Therefore, both $p$ and $q$ will next learn the decision of the instance $k+1$. From the agreement property of consensus, the $msgSet$ decided is the same for both processes. In the Algorithm \ref{algorithm:deliveryminimal}, the timestamps may be changed after the consensus decision only in line \ref{algline:changetsafterdecision}, based on what is already in the $\decided$ set (lines \ref{algline:decidedloopbegin} to \ref{algline:changetsafterdecision}). As the $msgSet$ and $\decided$ sets are each identical in $p$ and $q$, they will make the exact same change to the timestamp of each message in the \textit{while} loop (lines \ref{algline:decidedloopbegin} to \ref{algline:rmothers}). Therefore, they also agree on the timestamps of each message decided on consensus instance $k+1$ and their $\decided$ sets remain identical.\\

Now, regarding the case of $q$ not belonging to $g$, let $r$ be any correct process of such group. After setting the final timestamp of $m$ (l. \ref{algline:decidedloopbegin} to l. \ref{algline:rmothers}), $r$ \rmcast{}s $m$ to $group(q)$ (l. \ref{algline:rmothers}). From the algorithm, after \rmdel{}ing $m$, $q$ never changes the value of $m.ts$, set by $q$ in accordance with $p$.% As any two processes on the same group from which $m$ has been multicast agree on its timestamp, and since any process from other group which is also a destination of $m$ agree on its timestamp as well, than any two processes that are destinations of $m$ agree on its timestamp.
\end{proof}

@@

\begin{proof}
In Algorithm \ref{algorithm:deliveryminimal}, every process $q$ from $h$ sends messages to the processes of $g$, including $p$, using a fifo channel, via the \rmcast{} primitive. From the properties of consensus, we know that all correct processes from $h$ will decide the same messages in each consensus instance $k$. From Lemma \ref{lemma:lowertsfrmcastfirst} and Lemma \ref{lemma:agreetimestamps}, every correct process from $h$ \rmcast{}s to $g$ every message $m : m.src = h \wedge g \in m.dst$ (l. \ref{algline:rmothers} of Algorithm \ref{algorithm:deliveryminimal}) in the same order, which is that of the timestamps. From the fifo property of \rmcast{}, we know that $p$ will \rmdel{} all of such messages from each process of $h$ in ascending order of timestamps. Although such messages will arrive from different senders and may be out of order at $p$, once $p$ received some message $m$ from some process of $h$, $p$ has also received any message \mbox{$m' : m'.src = h \wedge g \in m.dst\wedge m'.ts < m.ts$}.
%
%Assume, by way of contradiction, that $p$ received a message $m' : m'.ts < m.ts$ from some process $r \in h$ after having received $m$ from $q$. As $m'.src = m.src$ and $q$ \rmcast{}s to $g$ every message $m : m.src = h \wedge g \in m.dst$, this means that $q$ will send $m'$ afterwards, contradicting the fact that the processes in $h$ agree on each batch of messages decided and send them to $g$ in the same fifo order.
\end{proof}

@@

\begin{proof}
Let $g$ be the group of $p$. Group $g$ is either $m.src$ or not. In the former case, before inserting any message $m' : m'.src = m.src$ into $\stamped$, $p$ has to decide $m'$ first. When a message is decided by $p$, $p$ checks whether some other message which was decided previously has a timestamp lower than $m'$ and, if that is the case, the timestamp of $m'$ is changed to a value greater than that of any other message already decided (lines \ref{algline:decidedloopbegin} to \ref{algline:changetsafterdecision} of Algorithm \ref{algorithm:deliveryminimal}). Only then $m'$ may be inserted into the $\stamped$ set of $p$ (l. \ref{algline:insbp2}). Therefore, $p$ inserts messages from other processes of $g$ into its $\stamped$ set in ascending order of timestamps. Besides, the messages from $g$ are \rmcast{} to other groups in this same order (l. \ref{algline:rmothers}) by all correct processes of $g$, from the uniform agreement property of consensus.

@@

%===           ====================================
%===   LEMMA   ====================================
%===           ====================================
\begin{lems} \label{lemma:oncebpwilldel}
Once a message $m \neq null$ has been inserted into the $\stamped$ set of a correct process $p$, $m$ is eventually delivered by $p$.
\end{lems}

\begin{proof}
Let $g = group(p)$. Eventually, $p$ will have received, from every group in $sendersTo(g)$, some message with a timestamp greater than $m.ts$ (from Lemma \ref{lemma:barrierperiodic}). Once this happens, no more messages with a timestamp lower than that of $m$ will be inserted into the $\stamped$ set of $p$ (from Lemma \ref{lemma:oncedecidednotsmaller} and the definition of $sendersTo(g)$). Let $ready$ be the set of $m$ plus all undelivered messages from $\stamped$ whose timestamps are lower than that of $m$, that is, $ready = \{m\} \cup \{m' : m' \in \stamped \setminus \delivered \wedge m'.ts < m.ts\}$.

We can infer that no more messages will be included in this set, meaning that any other message that might be later inserted into $\stamped$ will have a timestamp greater than that of any message in $ready$. Besides, the barriers received apply to all messages in this set, since their timestamps are not greater than $m.ts$. Each one of these messages will eventually be inserted into $\delivered$ and, if it is different from $null$, it will be delivered by $p$. We prove this by induction on the position $i$ of each message $m_i$ in $ready$, in ascending order of timestamps.\\
\\
\noindent Base case ($i=1$): Let $m_1$ be the first message in $ready$, i.e., \mbox{$\nexists m \in \stamped \setminus \delivered : m.ts < m_1.ts$}. We know that all the necessary barriers have been received already, so $m_1$ satisfies all conditions from lines \ref{algline:condel1} and \ref{algline:condel2} of \mbox{Algorithm \ref{algorithm:deliveryminimal}}. Therefore, $m_1$ will be inserted into $\delivered$, after which $m_1$ will no longer belong to $\stamped \setminus \delivered$. Also, if $m_1 \neq null$, $m_1$ will be delivered.\\
\\
\noindent Induction step: Suppose that $m_{i}$ will eventually be inserted into the $\delivered$ set. Once this happens, it means that $m_{i}$ was the first message in $\stamped \setminus \delivered$ in ascending timestamp order. Since no more messages have been inserted into $ready$, as soon as $m_{i}$ is inserted into $\delivered$, $m_{i+1}$ will be the first one in $\stamped \setminus \delivered$, having, therefore, the lowest timestamp in such set. Then, as all barriers necessary for $m$ have already been received and $m_{i+1}.ts \leq m.ts$, $m$ will satisfy both conditions of lines \ref{algline:condel1} and \ref{algline:condel2} of \mbox{Algorithm \ref{algorithm:deliveryminimal}}. Thus, $m_{i+1}$ will also be inserted into the $\delivered$ set and, if $m_{i+1} \neq null$, it will be delivered.
\end{proof}

@@

\begin{proof}
From lines \ref{algline:condel1} to \ref{algline:addtodelivered} of Algorithm \ref{algorithm:deliveryminimal}, and since no message is removed from $\delivered$, no message can be delivered twice, satisfying (3). For a message $m$ to be \amdel{}ed (l. \ref{algline:consdeliver}), it must belong to the $\stamped$ set (l. \ref{algline:condel1}).  There are two possibilities to when $m$ has been inserted into such set:
\begin{itemize}
  \item $m$ has been originated in the same same group of $p$, that is, $p \in m.src$, which means that $m$ was inserted into $\stamped$ in line \ref{algline:insbp2} of Algorithm \ref{algorithm:deliveryminimal}, or
  \item $m$ has been sent from a group $h \neq g$, which means that it was inserted by $p$ into $\stamped$ in line \ref{algline:insbp1}. 
\end{itemize}

@@

As for the case when $p \notin m.src$, $m$ is inserted into the $stamped$ set of process $p$ in line \ref{algline:insbp1}, when it has been \rmdel{}ed. For a process $q \in h$, where $h = m.src$, to \rmcast{} $m$, $m$ must have been decided within $h$, which means that it was proposed within $h$. Therefore it was \amcast{} -- satisfying condition (1) -- by $h$ to $g$. From line \ref{algline:rmothers}, $g$ necessarily belongs to $m.dst$, satisfying condition (2).

@@

\begin{proof}
Suppose, by way of contradiction, that $m'$ is delivered by $p$ before $m$. Message $m$ either has already been inserted into $\stamped$ or not. In the former case, as $m \neq null$ and $m$ has not been delivered, then $m$ belongs to $\stamped \setminus \delivered$. Therefore, $m'$ could not have been delivered, since $m.ts < m'.ts$ and it would not satisfy the condition from line \ref{algline:condel1} until $m$ has been delivered, so we have a contradiction in the case where $m$ was already in $\stamped$.

The other case is when $m$ has not been inserted into $\stamped$. From line \ref{algline:condel2} of Algorithm \ref{algorithm:deliveryminimal}, and since $m'$ has been delivered, we know that $p$ has received some barrier $b > m'.ts$ from every group in $sendersTo(group(p))$. Therefore, from Lemma \ref{lemma:oncedecidednotsmaller}, and since any barrier is also a message, we know that any new message that arrives at $p$ will have a timestamp greater than $m'.ts$. As $m$ has arrived after $m'$, then $m.ts > m'.ts$, which is also a contradiction.
